\pkg{cubble}: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2022-06-27 1-intro.Rmd
visualisation -> visualization

The SpatioTemporal CRAN task view [@ctvspatiotemporal] gathers information about R packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} [@spacetime] implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The \pkg{stars} [@stars] package has a new implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with \pkg{sf} [@sf], a package commonly used for wrangling spatial data, and the \pkg{tidyverse} [@tidyverse] suite for general data wrangling and visualisation in \proglang{R}.

@@

The other reason is about tidy data concepts [@tidydata] and how they should be applied to spatio-temporal data. According to the tidy data principles, data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data form given the downstream software such as \pkg{dplyr} [@dplyr] and \pkg{ggplot2} [@ggplot2] for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for hourly, daily or sub-daily periods over years, which are extensively collected and handled, for example in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make data wrangling, visualizing and analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full grid, sparse grid, or irregular, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at https://CRAN.R-project.org/package=cubble.

The rest of the paper is organized as follows: Section \@ref(cube) introduces the proposed cube structure as a way to conceptualise multivariate spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of \pkg{cubble}. Section \@ref(others) explains how cubble deals with more advanced considerations, including data with hierarchical structure, data matching and how cubble fits with existing static and interactive visualisation tools. Moreover we also illustrate how \pkg{cubble} deals with spatio-temporal data transformations. Section \@ref(examples) uses Australian weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \@ref(conclude) discuss the paper contributions and future directions.

>>>>

The SpatioTemporal CRAN task view [@ctvspatiotemporal] gathers information about R packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} [@spacetime] implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The \pkg{stars} [@stars] package has a new implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with \pkg{sf} [@sf], a package commonly used for wrangling spatial data, and the \pkg{tidyverse} [@tidyverse] suite for general data wrangling and visualization in \proglang{R}.

@@

The other reason is about tidy data concepts [@tidydata] and how they should be applied to spatio-temporal data. According to the tidy data principles, data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data form given the downstream software such as \pkg{dplyr} [@dplyr] and \pkg{ggplot2} [@ggplot2] for data wrangling and visualization. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for hourly, daily or sub-daily periods over years, which are extensively collected and handled, for example in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make data wrangling, visualizing and analyzing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organize spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronized. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full grid, sparse grid, or irregular, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at https://CRAN.R-project.org/package=cubble.

The rest of the paper is organized as follows: Section \@ref(cube) introduces the proposed cube structure as a way to conceptualize multivariate spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of \pkg{cubble}. Section \@ref(others) explains how cubble deals with more advanced considerations, including data with hierarchical structure, data matching and how cubble fits with existing static and interactive visualization tools. Moreover we also illustrate how \pkg{cubble} deals with spatio-temporal data transformations. Section \@ref(examples) uses Australian weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \@ref(conclude) discuss the paper contributions and future directions.

### 2022-06-27 2-cube.Rmd
visualisation -> visualization

Spatio-temporal data can be conceptualised using a cubical data model with three axes which typically are, time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation purposes: @lu_multidimensional_2018 shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; @bach_review_2014 reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but refers to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisations, still applies to our spatio-temporal data manipulation and visualisation approaches.

The most common space-time cube uses the three axes, time, latitude, longitude, and can be considered stacking space across time. Ours is a multivariate spatio-temporal cube with the three axes defined to be time, site and variables, as illustrated in the leftmost column of Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the conceptual framework behind the \pkg{cubble} objects. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.

>>>>

Spatio-temporal data can be conceptualized using a cubical data model with three axes which typically are, time, latitude and longitude. This abstraction can be useful for generalizing operations and visualization purposes: @lu_multidimensional_2018 shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; @bach_review_2014 reviews the temporal data visualization based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but refers to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualizations, still applies to our spatio-temporal data manipulation and visualization approaches.

The most common space-time cube uses the three axes, time, latitude, longitude, and can be considered stacking space across time. Ours is a multivariate spatio-temporal cube with the three axes defined to be time, site and variables, as illustrated in the leftmost column of Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hyper-cubes when describing multivariate spatio-temporal data and is the conceptual framework behind the \pkg{cubble} objects. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.

### 2022-06-27 3-the-cubble-pkg.Rmd
visualisation -> visualization

A \code{cubble} object is an S3 class [@advr] built on the \code{tibble} class, specifically to organise spatio-temporal data. The \code{cubble} class uses an attribute "form",  to arrange the spatial or temporal data components tidily. The form attribute can take a value of either "nested" or "long". The nested cubble is a subclass of rowwise tibble (\code{rowwise_df}). It arranges each spatial site in a row, and uses list columns to store the feature geometry and the temporal information. The long cubble is a subclass of grouped tibble (\code{grouped_df}), which expands the temporal information into the long form and stores the spatial information in a "spatial" attribute.

@@

NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention [@hassell_data_2017] to standardise the format of the attributes. A few packages in R exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

>>>>

A \code{cubble} object is an S3 class [@advr] built on the \code{tibble} class, specifically to organize spatio-temporal data. The \code{cubble} class uses an attribute "form",  to arrange the spatial or temporal data components tidily. The form attribute can take a value of either "nested" or "long". The nested cubble is a subclass of rowwise tibble (\code{rowwise_df}). It arranges each spatial site in a row, and uses list columns to store the feature geometry and the temporal information. The long cubble is a subclass of grouped tibble (\code{grouped_df}), which expands the temporal information into the long form and stores the spatial information in a "spatial" attribute.

@@

NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention [@hassell_data_2017] to standardize the format of the attributes. A few packages in R exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

### 2022-06-27 4-other-features.Rmd
visualisation -> visualization

Spatial locations can have grouping structures either inherent to the data (e.g. state within country) or obtained during the analysis (e.g. cluster id). In this case, it can be useful to summarise variables at various levels of the hierarchy. The function \code{switch_key()} can be used to change the grouping level of spatial locations. The diagram in Figure \ref{fig:illu-hier} shows how this function can be used to switch the grouping from station ids to cluster ids. The result can also be stretched into long form. By specifying

@@

Spatio-temporal data lends itself to a range of transformations. Glyph maps (Section \ref{unfold}) transform the measured variable and time coordinates into microplots at the spatial locations. Calendar plots [@wang2020calendar] deconstruct time to produce plots of variables in a calendar format. Summarising multiple variables is commonly done using projections, or linear combinations. Here we elaborate on the transformations made to produce a glyph map.

>>>>

Spatial locations can have grouping structures either inherent to the data (e.g. state within country) or obtained during the analysis (e.g. cluster id). In this case, it can be useful to summarize variables at various levels of the hierarchy. The function \code{switch_key()} can be used to change the grouping level of spatial locations. The diagram in Figure \ref{fig:illu-hier} shows how this function can be used to switch the grouping from station ids to cluster ids. The result can also be stretched into long form. By specifying

@@

Spatio-temporal data lends itself to a range of transformations. Glyph maps (Section \ref{unfold}) transform the measured variable and time coordinates into microplots at the spatial locations. Calendar plots [@wang2020calendar] deconstruct time to produce plots of variables in a calendar format. Summarizing multiple variables is commonly done using projections, or linear combinations. Here we elaborate on the transformations made to produce a glyph map.

### 2022-06-27 5-1-covid.Rmd
visualisation -> visualization

Since the start of the pandemic, the Victoria State Government in Australia has provided daily COVID counts by local government area (LGA). This data can be used to visualise COVID incidence and spread spatially, when combined with map polygon data available from the Australian Bureau of Statistics. These different sources need to be combined for the analysis, by matching the LGA names. Here is how to do this with \pkg{cubble}, including how to handle mismatches arising from different names of the same LGAs in the two tables. The COVID data is store in a `csv` file and looks like:

>>>>

Since the start of the pandemic, the Victoria State Government in Australia has provided daily COVID counts by local government area (LGA). This data can be used to visualize COVID incidence and spread spatially, when combined with map polygon data available from the Australian Bureau of Statistics. These different sources need to be combined for the analysis, by matching the LGA names. Here is how to do this with \pkg{cubble}, including how to handle mismatches arising from different names of the same LGAs in the two tables. The COVID data is store in a `csv` file and looks like:

### 2022-06-27 5-3-river-level-rainfall.Rmd
visualisation -> visualization

This function returns a \code{cubble} object, with additional columns: \code{dist} storing the distance between matched stations, \code{group} summarising spatial matching, and \code{n_match} showing the temporal matching.

@@

Figure \ref{fig:matching} shows four matched pairs on the map (a) and standardised data as time series (b). The expected concurrent increase in precipitation and water level can be seen clearly.

>>>>

This function returns a \code{cubble} object, with additional columns: \code{dist} storing the distance between matched stations, \code{group} summarizing spatial matching, and \code{n_match} showing the temporal matching.

@@

Figure \ref{fig:matching} shows four matched pairs on the map (a) and standardized data as time series (b). The expected concurrent increase in precipitation and water level can be seen clearly.

### 2022-06-27 5-4-era5.Rmd
visualisation -> visualization

<!--The goal of this example is to reproduce the break-up of the southern polar vortex that happened in late September and early October in 2002 in the stratosphere in the Southern Hemisphere. Therefore, we filter data corresponding to the Southern Hemisphere and restrict our analysis to four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04.--> The downloaded netcdf data is read into R using the \code{ncdf4} package, and converted to a \code{cubble}:

>>>>

<!--The goal of this example is to reproduce the break-up of the southern polar vortex that happened in late September and early October in 2002 in the stratosphere in the Southern Hemisphere. Therefore, we filter data corresponding to the Southern Hemisphere and restrict our analysis to four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04.--> The downloaded netCDF data is read into R using the \code{ncdf4} package, and converted to a \code{cubble}:

### 2022-06-27 5-5-interactive-graphics.Rmd
visualisation -> visualization

The key steps are to convert both the nested and long forms of the data into shared objects, and to plot these sid-by-side. The two are linked by the station identifier.

@@

Plot (a) shows the initial state of the interactive display: all locations are shown as dots on the map, coloured by temperature range, and the right plot shows the ribbons representing maximum to minimum for all stations. In plot (b) the "mount elizabeth" station, which shows a high variance colour on the initial map, is selected on the map and this produces the ribbon on the right. In plot (c) the lowest temperature in August is selected, which is "thredbo" station on the left map. It was surprising to us that this was not a station in Tasmania, so for comparison a station in Tasmania is selected on the map to show in relation to thredbo. We can see that thredbo has a bigger winter dip in temperature, and although Tasmania is cold generally, it's temperatures are more consistent.

>>>>

The key steps are to convert both the nested and long forms of the data into shared objects, and to plot these side-by-side. The two are linked by the station identifier.

@@

Plot (a) shows the initial state of the interactive display: all locations are shown as dots on the map, coloured by temperature range, and the right plot shows the ribbons representing maximum to minimum for all stations. In plot (b) the "Mount Elizabeth" station, which shows a high variance colour on the initial map, is selected on the map and this produces the ribbon on the right. In plot (c) the lowest temperature in August is selected, which is "Thredbo" station on the left map. It was surprising to us that this was not a station in Tasmania, so for comparison a station in Tasmania is selected on the map to show in relation to Thredbo. We can see that Thredbo has a bigger winter dip in temperature, and although Tasmania is cold generally, it's temperatures are more consistent.

### 2022-06-27 6-conclusion.Rmd
misc small fixes
visualisation -> visualization

This paper presents an \proglang{R} package \pkg{cubble} for organising, manipulating and visualising spatio-temporal data. The package introduces a new data structure for spatio-temporal data, \code{cubble}, that connects the time invariant and varying variables and that allows the user to work with a nested and long form of the data. This work adds capabilities into the spatio-temporal practitioners toolbox to integrate it with a tidy data framework. The data structure and functions introduced in this package can be used and combined with existing spatio-temporal R packages such as \pkg{sf}, data wrangling packages such as \pkg{dplyr}, and visualization packages such as \pkg{ggplot2}, \pkg{plotly} and \pkg{leaflet}.

Numerous examples are provided in the main text, appendix and package vignettes. These include creating and coercing data with mismatched names, handling hierarchical data, matching time series spatially and temporally, reproducing ERA5 plots from NetCDF data. Visualization of the \code{cubble} objects can be done with interactive graphic pipelines using \pkg{crosstalk}, \pkg{plotly} and \pkg{leaflet}.

@@

This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship and started while Nicolas Langrené was affiliated with CSIRO's Data61. The article is created using \pkg{knitr} [@knitr] and \pkg{rmarkdown} [@rmarkdown] in R. The source code for reproducing this paper can be found at: https://github.com/huizezhang-sherry/paper-cubble.

>>>>

This paper presents an \proglang{R} package \pkg{cubble} for organizing, manipulating and visualizing spatio-temporal data. The package introduces a new data structure for spatio-temporal data, \code{cubble}, that connects the time invariant and varying variables and that allows the user to work with a nested and long form of the data. This work adds capabilities into the spatio-temporal practitioners toolbox to integrate it with a tidy data framework. The data structure and functions introduced in this package can be used and combined with existing spatio-temporal R packages such as \pkg{sf}, data wrangling packages such as \pkg{dplyr}, and visualization packages such as \pkg{ggplot2}, \pkg{plotly} and \pkg{leaflet}.

Numerous examples are provided in the main text, appendix and package vignettes. These include creating and coercing data with mismatched names, handling hierarchical data, matching time series spatially and temporally, reproducing ERA5 plots from netCDF data. Visualization of the \code{cubble} objects can be done with interactive graphic pipelines using \pkg{crosstalk}, \pkg{plotly} and \pkg{leaflet}.

@@

This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship and started while Nicolas Langrené was affiliated with CSIRO's Data61. The article is created using \pkg{knitr} [@knitr] and \pkg{rmarkdown} [@rmarkdown] in R using the `rticles::jss_article` template. The source code for reproducing this paper can be found at: https://github.com/huizezhang-sherry/paper-cubble.

### 2022-06-27 jss/main.Rmd
visualisation -> visualization

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: 'Monash University'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
    Multivariate spatio-temporal data refers to multiple measurements taken across space and time. For many analyses, spatial and time components can be separately studied: for example, to explore the temporal trend of one variable for a single spatial location, or to model the spatial distribution of one variable at a given time. However for some studies, it is important to analyse different aspects of the spatio-temporal data simultaneouly, like for instance, temporal trends of multiple variables across locations. In order to facilitate the study of different portions or combinations of spatio-temporal data, we introduce a new data structure, \code{cubble}, with a suite of functions enabling easy slicing and dicing on the different components spatio-temporal components.  The proposed \code{cubble} structure ensures that all the components of the data are easy to access and manipulate while providing flexibility for data analysis. In addition, \code{cubble} facilitates visual and numerical explorations of the data while easing data wrangling and modelling. The \code{cubble} structure and the functions provided in the \code{cubble} R package equip users with the capability to handle hierarchical spatial and temporal structures. The \code{cubble} structure and the tools implemented in the package are illustrated with different examples of Australian climate data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
  \newcommand{\sherry}[1]{\textcolor{magenta}{#1}}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
bibliography: ../references.bib
---

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: 'Monash University'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
     Multivariate spatio-temporal data refers to multiple measurements taken across space and time. For many analyses, spatial and time components can be separately studied: for example, to explore the temporal trend of one variable for a single spatial location, or to model the spatial distribution of one variable at a given time. However for some studies, it is important to analyze different aspects of the spatio-temporal data simultaneously, like for instance, temporal trends of multiple variables across locations. In order to facilitate the study of different portions or combinations of spatio-temporal data, we introduce a new data structure, \code{cubble}, with a suite of functions enabling easy slicing and dicing on the different components spatio-temporal components.  The proposed \code{cubble} structure ensures that all the components of the data are easy to access and manipulate while providing flexibility for data analysis. In addition, \code{cubble} facilitates visual and numerical explorations of the data while easing data wrangling and modelling. The \code{cubble} structure and the functions provided in the \code{cubble} R package equip users with the capability to handle hierarchical spatial and temporal structures. The \code{cubble} structure and the tools implemented in the package are illustrated with different examples of Australian climate data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
  \newcommand{\sherry}[1]{\textcolor{magenta}{#1}}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
bibliography: ../references.bib
---
